---
title: 'Project 2: Modeling, Testing, and Predicting'
author: "Daelon Morais djm4445"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(echo = TRUE, eval = TRUE,fig.align="center",warning=FALSE,message=FALSE,fig.width=8, fig.height=5, linewidth=60)
options(tibble.width = 100,width = 100)
library(dplyr);library(ggplot2);library(tidyverse);library(readxl);library(lmtest);library(sandwich)
```

## Data
```{r}
incomePoverty <- read_excel("~/Documents/UT/Classes/SDS 348/incomePoverty.xlsx")
educationEmployment <- read_excel("~/Documents/UT/Classes/SDS 348/educationEmployment.xlsx")
opioidEDvisits <- read_excel("~/Documents/UT/Classes/SDS 348/opioidEDvisits.xlsx")
fulldata <- opioidEDvisits %>% full_join(incomePoverty, by="County")%>% full_join(educationEmployment, by="County")%>% na.omit(fulldata)
```
  The first dataset contains income and poverty data by county for all counties in Texas. It contains the variables: Counties, Per Capita Income, Median Household Income, and % of Population in Poverty. The second dataset contains data about the level of education and employment by county and contains the variables: Counties, % High School Graduate or Higher, % Bachelor s Degree or Higher, and Unemployment Rate (%). The third dataset contains data on opioid-related emergency department visits in 2017 by county. It contains the variables: Counties and Number of Opioid-Related Emergency Department Visits. Some data is suppressed from some of the counties due to privacy concerns and patient protection.
  The first two datasets cam from the same source, the Texas Association of Counties. You can select a variety of variables and export the data from their website. The third dataset regard opioid-related visits to the ED came from the Department of State Health Services. They have an interactive data visualizer through their website, but I had to email them to get the raw CSV file.  
  I chose these datasets becuase they are similar to the project I did in BIO328M but also very different. Back then, and currently, I am interest in emergency medicine as a potential career path. In biostatics, I looked into the association between education level and the number of trauma cases by county. Since that time, I have begun to be interested in the ongoing opioid epidemic and have decide to pursue research regarding the precribing of opioid in the emergency department. Although income, poverty, education, and employment are not variables I am investigating in my thesis, I thought it would be interesting to look into any possible associations between the variables.  
  I assume to see poverty levels and unemployment rates to bepositively correlated to the number of opioid-related visits to the ED while expecting income and education to be negatively correlated to the number of opioid-related visits to the ED. This is due to socio-economic variables that affect drug abuse. The opioid data can be due to improper discharge instructions, lack of education about the dangers of the medication, addiction, and more. 

## MANOVA
```{r}
#create a categorical response variable, split into 3, 1=low, 2=medium, 3=high
fulldata$EDvisits_cat <- as.numeric(cut_number(fulldata$`Number of Opioid-Related Emergency Department Visits`,3))
#manova test
man1<-manova(cbind(`% Bachelor s Degree or Higher`,`% High School Graduate or Higher`,`% of Population in Poverty`,`Unemployment Rate (%)`,`Median Household Income`, `Per Capita Income`)~EDvisits_cat, data=fulldata)
summary(man1)
#univariate ANOVA test
summary.aov(man1)
#post-hoc t tests
pairwise.t.test(fulldata$`% High School Graduate or Higher`,fulldata$EDvisits_cat,p.adj="none")
pairwise.t.test(fulldata$`% Bachelor s Degree or Higher`,fulldata$EDvisits_cat,p.adj="none")
pairwise.t.test(fulldata$`Median Household Income`,fulldata$EDvisits_cat,p.adj="none")
#Bonferroni Adjustment: at least 1 type I error
1-(.95)^16
.05/16
```
For the MANOVA, the hypotheses are: H0: For `% Bachelor s Degree or Higher`,`% High School Graduate or Higher`,`% of Population in Poverty`,`Unemployment Rate (%)`,`Median Household Income`, `Per Capita Income`, means of each level of EDvisits are equal; Ha:  For at least one DV, at least one EDvisits level mean is different. A one-way multivariate analysis of variance (MANOVA) was conducted to determine the effect of the iris Species (setosa, versicolor, virginica) on two dependent variables (`% Bachelor s Degree or Higher`,`% High School Graduate or Higher`,`% of Population in Poverty`,`Unemployment Rate (%)`,`Median Household Income`, and `Per Capita Income`).  Significant differences were found among the dependent variables. Pillai trace=0.23287, pseudo F(120,6)=,p<.0001)
Univariate analyses of variance (ANOVAs) for each dependent variable were conducted as follow-up tests to the MANOVA, using the Bonferroni method for controlling Type I error rates for multiple comparisons. The univariate ANOVAs for the three EDvisit categories on the  dependent
measures % Bachelor s Degree or Higher (p<5.08e-06), % High School Graduate or Higher (p<0.002376), and Median Household Income (p<0.004052).Since I have 1 MANOVA with 3 significant numeric response variables, I have have 1 (MANOVA) + 6 (ANOVA) + 3*3 (t-tests) for a total of 10 tests. So probability of at least one type I error is 0.5598733 My new p-value is .003.
Post hoc analysis was performed conducting pairwise comparisons to determine which ED visit categories differed in % Bachelor s Degree or Higher, % High School Graduate or Higher, and Median Household Income. For % High School Graduate or Higher, the lowest and highest categories of the number of ED visits significantly differed (p=0.0026). % Bachelor s Degree or Higher significantly differed between the middle group and the highest group (p=2.3e-06) and the lowest group and the highest group((p=1.7e-06). Median Household Income did not significantly differ among the groups . 
The univariate anova tests confirmed the differences. Regarding the assumptions, this is not a random sample since it is all the counties in Texas. Additionally, the dependent variables are probably correlated and most likely have a linear relatioship since education and income are typically associated which violates another assumption. There may be some extreme valus since some counties have very samll populations. The other assumptions of Multivariate normality of DVs and  Homogeneity of within-group covariance matrices might be met. 

## Radomization Test
```{r}
#randomization test
obs_F<-22.74
Fs<-replicate(5000,{
 new<-fulldata%>%mutate(Bachelors_sample=sample(`% Bachelor s Degree or Higher`))
 SSW<- new%>%group_by(EDvisits_cat)%>%summarize(SSW=sum((Bachelors_sample-mean(Bachelors_sample))^2))%>%summarize(sum(SSW))%>%pull
 SSB<- new%>%mutate(mean=mean(Bachelors_sample))%>%group_by(EDvisits_cat)%>%mutate(groupmean=mean(Bachelors_sample))%>%
 summarize(SSB=sum((mean-groupmean)^2))%>%summarize(sum(SSB))%>%pull
 (SSB/2)/(SSW/57)
})
#plot visualizing the null distribution and the test statistic
hist(Fs, prob=T); abline(v = obs_F, col="red",add=T)
mean(Fs>obs_F)
```

The null hypothesis is H0: for each ED visit level, the means of the groups of ED visits are equal and Ha: For at least one ED visit category, at least one group of ED visits mean differs. Since the mean(Fs>obs_F) is 0, we reject the null and say that the % of the counties population with Bachelor s Degree or Higher differs across varying levels of ED visits. 

## Linear Regression Model
```{r}
EDvisits<-fulldata$`Number of Opioid-Related Emergency Department Visits`#-mean(fulldata$`Number of Opioid-Related Emergency Department Visits`)
Bachelors_c<-fulldata$`% Bachelor s Degree or Higher`-mean(fulldata$`% Bachelor s Degree or Higher`)
MedIncome_c<-fulldata$`Median Household Income`-mean(fulldata$`Median Household Income`)
MedIncome<-fulldata$`Median Household Income`
centeredData<-data.frame(EDvisits,Bachelors_c, MedIncome_c,MedIncome)
fit<-lm(EDvisits ~MedIncome_c*Bachelors_c, data=centeredData)
summary(fit)
#Plot of the Regression
new1<-centeredData
new1$MedIncome_c<-mean(centeredData$MedIncome_c)
new1$mean<-predict(fit,new1)
new1$MedIncome_c<-mean(centeredData$MedIncome_c)+sd(centeredData$MedIncome_c)
new1$plus.sd<-predict(fit,new1)
new1$MedIncome_c<-mean(centeredData$MedIncome_c)-sd(centeredData$MedIncome_c)
new1$minus.sd<-predict(fit,new1)
newint<-new1%>%dplyr::select(EDvisits,Bachelors_c,mean,plus.sd,minus.sd)%>%gather(MedIncome,value,-EDvisits,-Bachelors_c)

mycols<-c("#619CFF","#F8766D","#00BA38")
names(mycols)<-c("-1 sd","mean","+1 sd")
mycols=as.factor(mycols)

ggplot(centeredData,aes(Bachelors_c,EDvisits),group=mycols)+geom_point()+geom_line(data=new1,aes(y=mean,color="mean"))+geom_line(data=new1,aes(y=plus.sd,color="+1 sd"))+geom_line(data=new1,aes(y=minus.sd,color="-1 sd"))+scale_color_manual(values=mycols)+labs(color="Median Income ($) (cont)")+theme(legend.position=c(.9,.2))

#Check Assumptions
plot(fit)
resids<-fit$residuals
#homoskedasticity 
bptest(fit) 
#linearity
shapiro.test(resids) #Ho: distribution is normal
#normality
ks.test(resids, "pnorm", mean=0, sd(resids)) #Ho: distribution is normal

#corrected SE
coeftest(fit, vcov = vcovHC(fit))

#main effects
fit1<-lm(EDvisits ~MedIncome_c+Bachelors_c, data=centeredData)
summary(fit1)
```
The Best-fitting line: # of ED Visits = 6.700e+01 + 6.549e+00(% with Bachelors Degree). % with Bachelors Degree explains a significant amount of variation in Opioid ED Visits (t=2.599, df =123, p<.05 ). 
When median income and % with a Bachelor's Degree or higher are average, there are 67.6692913 ED visits on average. While holding median income constant, for every one percent average increase in % with Bachelors Degree, there is a 6.549 average increase in visits to the ED. While holding % of Bachelors degree constant, on average, every $1 decrease in median income results in a -3.812e-04 in # of ED visits. There's no significant difference in slope for ED Visits v % Bachelor's degree  based on varying levels of median income. 
The model explains 0.09122 of the variation in the Number of ED visits.
The BP test (P>.05) means that my data is homoskedastic, Shapiro-Wilk normality (p<.001) means my data is not linear, and the Kolmogorov-Smirnov test (p<.001) means the data is not normal.
With the Robust SEs, the estimates did not change. The SEs increased for the intercept and was significant (p<.001). The estimate stayed the same for the Median Income, decreased greatly for the % of Bahelor's Degrees or higher, and decreased for the interaction between median income and % of Bachelor's degrees but none of these changes were significant.
## Linear Regression Model with Bootstrapped SEs
```{r}
set.seed(1234)
samp_distn<-replicate(5000, {
 boot_dat<-centeredData[sample(nrow(centeredData),replace=TRUE),]
 fit<-lm(EDvisits ~MedIncome_c*Bachelors_c, data=boot_dat)
 coef(fit)
})
samp_distn%>%t%>%as.data.frame%>%summarize_all(sd)
```
The estimates for SEs compared to the original SEs increased dramatically to 17.03567 for the intercept, Median income increased as well to 0.0007205464, % with Bachelor's degrees increased to 3.703015, and the interaction increased to 0.0001183407. Compared to the robust SEs: intercept increased, median income decreased, increased for the % with a Bachelor's degree or higher and decreased for the interaction. The p values increased for the robust SE median income and decreased for the Bachelor's degree.  
## Logistic Regression
```{r}
class_diag<-function(probs,truth){
 tab<-table(factor(probs>.5,levels=c("FALSE","TRUE")),truth)
 acc=sum(diag(tab))/sum(tab)
 sens=tab[2,2]/colSums(tab)[2]
 spec=tab[1,1]/colSums(tab)[1]
 ppv=tab[2,2]/rowSums(tab)[2]
 if(is.numeric(truth)==FALSE & is.logical(truth)==FALSE) truth<-as.numeric(truth)-1
 #CALCULATE EXACT AUC
 ord<-order(probs, decreasing=TRUE)
 probs <- probs[ord]; truth <- truth[ord]
 TPR=cumsum(truth)/max(1,sum(truth))
 FPR=cumsum(!truth)/max(1,sum(!truth))
 dup<-c(probs[-1]>=probs[-length(probs)], FALSE)
 TPR<-c(0,TPR[!dup],1); FPR<-c(0,FPR[!dup],1)
 n <- length(TPR)
 auc<- sum( ((TPR[-1]+TPR[-n])/2) * (FPR[-1]-FPR[-n]) )
 data.frame(acc,sens,spec,ppv,auc)
} 
lotsOfEDvisits<-ifelse(fulldata$`Number of Opioid-Related Emergency Department Visits`>median(fulldata$`Number of Opioid-Related Emergency Department Visits`),1,0)
#EDvisits<-fulldata$`Number of Opioid-Related Emergency Department Visits`
Bachelors<-fulldata$`% Bachelor s Degree or Higher`
MedIncome<-fulldata$`Median Household Income`
binaryData<-data.frame(lotsOfEDvisits,Bachelors, MedIncome)
fit<-glm(lotsOfEDvisits~Bachelors+MedIncome,data=binaryData,family=binomial(link="logit"))
coeftest(fit)
exp(coef(fit))
probs<-predict(fit,type="response")
class_diag(probs,binaryData$lotsOfEDvisits)
yhat<-predict(fit) 
mean((binaryData$lotsOfEDvisits-yhat)^2) #mean squared error (MSE)
#Confusion Martrix
pred<-ifelse(probs>.5,1,0) 
table(predictions=pred,truth=binaryData$lotsOfEDvisits)%>%addmargins

#plot density of log-odds (logit) by binary outcome variable
data<-binaryData
data$lotsOfEDvisits<-as.factor(data$lotsOfEDvisits)
data$logit<-predict(fit,type="link")

data%>%ggplot()+geom_density(aes(logit,color=lotsOfEDvisits,fill=lotsOfEDvisits), alpha=.4)+
  theme(legend.position=c(.85,.85))+geom_vline(xintercept=0)+xlab("predictor (logit)")

#ROC Curve
data$prob<-predict(fit,type="response") 
sens<-function(p,data=data, y=lotsOfEDvisits) mean(data[data$lotsOfEDvisits==1,]$prob>p)
spec<-function(p,data=data, y=lotsOfEDvisits) mean(data[data$lotsOfEDvisits==0,]$prob<p)
sensitivity<-sapply(seq(0,1,.01),sens,data)
specificity<-sapply(seq(0,1,.01),spec,data)
ROC1<-data.frame(sensitivity,specificity,cutoff=seq(0,1,.01))
ROC1$TPR<-sensitivity
ROC1$FPR<-1-specificity
ROC1%>%ggplot(aes(FPR,TPR))+geom_path(size=1.5)+geom_segment(aes(x=0,y=0,xend=1,yend=1),lty=2)+
 scale_x_continuous(limits = c(0,1))

#AUC
widths<-diff(ROC1$FPR) #horizontal distances
heights<-vector() #sum of top and bottom vertical line in loop
for(i in 1:100) heights[i]<-ROC1$TPR[i]+ROC1$TPR[i+1]
AUC<-sum(heights*widths/2) #sum up areas of trapezoids
AUC%>%round(3)

#10-fold
set.seed(1234)
k=10
data1<-binaryData[sample(nrow(binaryData)),] #put dataset in random order
folds<-cut(seq(1:nrow(binaryData)),breaks=k,labels=F) #create folds
diags<-NULL
for(i in 1:k){ # FOR EACH OF 10 FOLDS
  train<-data1[folds!=i,] #CREATE TRAINING SET
  test<-data1[folds==i,] #CREATE TESTING SET
  truth<-test$lotsOfEDvisits
  
  fit<- glm(lotsOfEDvisits~(.),data=binaryData,family="binomial")
  prob<- predict(fit,newdata = test,type="response")
  
  diags<-rbind(diags,class_diag(prob,truth)) 
}
apply(diags,2,mean)

```
'lotsOfED' is a binary variable I created that separated the Numer of opioid related ED visits into a 1,0's to represent a high or low number of visits. For $0 median income and 0% bachelors degree, the odds of having a high number of ED visits is 0.0622286. Every one-unit increase in % Bachelor's degrees or higher multiplies odds of having a high number of ED visits by 1.105281. Every one-unit increase in Median Income multiplies odds of having a high number of ED visits by 1.000014. However, Median income is not significant while  % Bachelor's  degree or higher is (p<.01). The accuracy is 0.6141732 which is the proportion of correctly classified level of ED visit counties. The sensitivity (TPR) is .5 which is the proportion of correctly identified high number ED visit counties. The specificity is 0.7230769 is the proportion of low number ED visits correctly classified. The PPV (precision) is 0.6326531 is the proportion of high number of ED visits correctly classified out of the high level ED visits. The AUC is 0.6858561 which is not a very good predictor because it is the probability that a randomly county that has a higher number of ED visits has a higher prediction than a randomly selected county with a low amount of ED visits. . After the 10 fold, The accuracy is 0.6141026, the sensitivity (TPR) is 0.4988095, the specificity is 0.7374603 is the proportion of high number ED visits correctly classified. The PPV (TNR) is 0.6469048  is proportion of benigns correctly classified. The AUC is 0.6776984 which is still not a very good predictor. The MSE is 0.9720963.

## Lasso Regression

- **6. (10 pts)** Choose one variable you want to predict (can be one you used from before; either binary or continuous) and run a LASSO regression inputting all the rest of your variables as predictors. Choose lambda to give the simplest model whose accuracy is near that of the best (i.e., `lambda.1se`). Discuss which variables are retained. Perform 10-fold CV using this model: if response in binary, compare model's out-of-sample accuracy to that of your logistic regression in part 5; if response is numeric, compare the residual standard error (at the bottom of the summary output, aka RMSE): lower is better fit!
```{r}
library(glmnet)
fulldata$lotsOfEDvisits<-ifelse(fulldata$`Number of Opioid-Related Emergency Department Visits`>median(fulldata$`Number of Opioid-Related Emergency Department Visits`),1,0)
y<-as.matrix(fulldata$lotsOfEDvisits)
x<-fulldata%>%dplyr::select(-c(County,`Number of Opioid-Related Emergency Department Visits`,lotsOfEDvisits))%>%mutate_all(scale)%>%as.matrix
cv<-cv.glmnet(x,y) 
lasso1<-glmnet(x,y,lambda=cv$lambda.1se)
coef(lasso1)
set.seed(1234)
k=10 #choose number of folds
data1<-fulldata[sample(nrow(fulldata)),] #randomly order rows
folds<-cut(seq(1:nrow(fulldata)),breaks=k,labels=F) #create folds
diags<-NULL
for(i in 1:k){
 train<-data1[folds!=i,]
 test<-data1[folds==i,]
 truth<-test$lotsOfEDvisits
 fit<-glm(lotsOfEDvisits~`% Bachelor s Degree or Higher`,data=train,family="binomial")
 probs<-predict(fit,newdata = test,type="response")
 preds<-ifelse(probs>.5,1,0)
 diags<-rbind(diags,class_diag(probs,truth))
}
diags%>%summarize_all(mean)
```
The % Bachelor s Degree or Higher variables was retained which means it is an important  and accurate predictor (0.4881890) for whether a county has a high number of opioid related ED visits or not. The model's out-of-sample accuracy is 0.6134615 which is less than the accuracy from part 5 which was 0.6141732 making the logistic regression a better model.
...





