---
title: "Project1 SDS348"
author: "Daelon Morais"
date: "10/13/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

  The first dataset contains income and poverty data by county for all counties in Texas. It contains the variables: Counties, Per Capita Income, Median Household Income, and % of Population in Poverty. The second dataset contains data about the level of education and employment by county and contains the variables: Counties, % High School Graduate or Higher, % Bachelor s Degree or Higher, and Unemployment Rate (%). The third dataset contains data on opioid-related emergency department visits in 2017 by county. It contains the variables: Counties and Number of Opioid-Related Emergency Department Visits. Some data is suppressed from some of the counties due to privacy concerns and patient protection.
  The first two datasets cam from the same source, the Texas Association of Counties. You can select a variety of variables and export the data from their website. The third dataset regard opioid-related visits to the ED came from the Department of State Health Services. They have an interactive data visualizer through their website, but I had to email them to get the raw CSV file.  
  I chose these datasets becuase they are similar to the project I did in BIO328M but also very different. Back then, and currently, I am interest in emergency medicine as a potential career path. In biostatics, I looked into the association between education level and the number of trauma cases by county. Since that time, I have begun to be interested in the ongoing opioid epidemic and have decide to pursue research regarding the precribing of opioid in the emergency department. Although income, poverty, education, and employment are not variables I am investigating in my thesis, I thought it would be interesting to look into any possible associations between the variables.  
  I assume to see poverty levels and unemployment rates to bepositively correlated to the number of opioid-related visits to the ED while expecting income and education to be negatively correlated to the number of opioid-related visits to the ED. This is due to socio-economic variables that affect drug abuse. The opioid data can be due to improper discharge instructions, lack of education about the dangers of the medication, addiction, and more. 

## 1. Tidying: Spread/Gather
```{r libraries}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(readxl)
incomePoverty <- read_excel("~/Documents/UT/Classes/SDS 348/incomePoverty.xlsx")
educationEmployment <- read_excel("~/Documents/UT/Classes/SDS 348/educationEmployment.xlsx")
opioidEDvisits <- read_excel("~/Documents/UT/Classes/SDS 348/opioidEDvisits.xlsx")
head(incomePoverty)
head(educationEmployment)
head(opioidEDvisits)
```
```{r Longer/Wider}
incomePovertyUntidy<-incomePoverty %>% pivot_longer(contains("Income"))
incomePovertyUntidy%>%head()
incomePovertyUntidy<-incomePovertyUntidy %>% pivot_wider(names_from="name",values_from="value")
incomePovertyUntidy%>%head()
educationEmploymentUntidy<-educationEmployment %>% pivot_longer(contains("%"))
educationEmploymentUntidy%>%head()
educationEmploymentUntidy<-educationEmploymentUntidy %>% pivot_wider(names_from="name",values_from="value")
educationEmploymentUntidy%>%head()     
```
All of the datasets were already tidy. The opioidEDvisits datasets only had one column of a dependent variable. The incomePoverty and educationEmployment data had more variables to work with. 
For the incomePoverty dataset, two of the columns had to do with income as a dollar amount while the last column concerned the poverty rate as a percentage. For that reason, I chose to make the dataset untidy by using pivot_longer to combine the income level columns which created a "name" column for per capita and median income level and a "value" column for the dollar amounts. I then used pivot_wider to separate the name column based on the name and the values followed, turning the dataset back to it's original form.

A similar process was carried out on the educationEmployment dataset; the three columns had to do with education level as a percentage of the population of the county. Therefore, I chose to make the dataset untidy by using pivot_longer to combine the columns that contained the "%" character which created a "name" column for each level of education and a "value" column for the percentage of the county's population that fit in each category. Then, I used pivot_wider to separate the name column based on the categories and the perccentages from the value column followed, returning the dataset to it's original form.

## 2. Joining/Merging
```{r Joining}
fulldata <- opioidEDvisits %>% full_join(incomePoverty, by="County")%>% full_join(educationEmployment, by="County")
before <-nrow(fulldata)
fulldata<- na.omit(fulldata)
after<- nrow(fulldata)
before-after
```
Since all of the datasets have the same counties as a common variable, I decided to use full_join. Using either left_join or right_join would have resulted in the same ending dataset and since there are matching columns in each, there are no rows dropped. 127 counties don't have data since it is suppressed due to privacy concerns and patient protection. I used na.omit() to remove these rows.


## 3. Wrangling
```{r Wrangling}
#generate new var using mutate:
fulldata %>% mutate(ED_visit_prop =`Number of Opioid-Related Emergency Department Visits`/sum(`Number of Opioid-Related Emergency Department Visits`)) %>% select(County,ED_visit_prop) %>%glimpse()
#generate categories based on % of population with a Bachelor's Degree or higher and per capita income
twoCatData<- fulldata %>% mutate(percentDegreeRange=cut(`% Bachelor s Degree or Higher`, breaks=c(-Inf,16.2,21.7,Inf), labels=c("low","middle","high")))%>% mutate(perCapitaIncomeRange=cut(`Per Capita Income`, breaks=c(-Inf,37809,43412,Inf), labels=c("low","middle","high")))
#1)mean ED visits by per capita income
twoCatData %>% select(c(perCapitaIncomeRange,`Number of Opioid-Related Emergency Department Visits`)) %>% group_by(perCapitaIncomeRange) %>%  summarize(mean_ED=mean(`Number of Opioid-Related Emergency Department Visits`,na.rm=T))
#2)mean % Bachelor's Degree or higher by per capita income
twoCatData %>% group_by(perCapitaIncomeRange) %>%  summarize(mean_Degree=mean(`% Bachelor s Degree or Higher`,na.rm=T))
#3)mean % High School Graduate or higher by per capita income
twoCatData %>% group_by(perCapitaIncomeRange) %>%  summarize(mean_High=mean(`% High School Graduate or Higher`,na.rm=T))
#4)mean Unemployment by per capita income
twoCatData %>% group_by(perCapitaIncomeRange) %>%  summarize(mean_Unemploy=mean(`Unemployment Rate (%)`,na.rm=T))
#5)mean % of Population in Poverty by per capita income
twoCatData %>% group_by(perCapitaIncomeRange) %>%  summarize(mean_Pov=mean(`% of Population in Poverty`,na.rm=T))

#6)mean per capita income by % of population with a Bachelor's Degree or higher
twoCatData %>% group_by(percentDegreeRange) %>%  summarize(mean_PerInc=mean(`Per Capita Income`,na.rm=T))
#7)mean median income by % of population with a Bachelor's Degree or higher
twoCatData %>% group_by(percentDegreeRange) %>%  summarize(mean_MedInc=mean(`Median Household Income`,na.rm=T))
#8)mean Unemployment by % of population with a Bachelor's Degree or higher
twoCatData %>% group_by(percentDegreeRange) %>%  summarize(mean_unemploy=mean(`Unemployment Rate (%)`,na.rm=T))
#9)mean % of Population in Poverty by % of population with a Bachelor's Degree or higher
twoCatData %>% group_by(percentDegreeRange) %>%  summarize(mean_Pov=mean(`% of Population in Poverty`,na.rm=T))
#10)mean ED visits by per capita income
twoCatData %>% group_by(perCapitaIncomeRange,percentDegreeRange) %>%  summarize(mean_ED=mean(`Number of Opioid-Related Emergency Department Visits`,na.rm=T)) %>% arrange(mean_ED)
#Of counties with high percentages of the population with a bachelor's degree or higher, what is the average poverty rate by per capita income
twoCatData%>% filter(percentDegreeRange=="high") %>%group_by(perCapitaIncomeRange) %>%  summarize(avg_Pov=mean(`% of Population in Poverty`)) 
```
I used mutate to generate the variable ED_visit_prop which is a function of the Numer of Opioid Related ED Visits; it is the proportion of Opioid ED Visits by County to the whole state of Texas. I then generated high, medium, and low categories based on % of population with a Bachelor's Degree or higher and per capita income in order to carry out the summary statistics. I calculated the mean ED visits by per capita income: low income counties had an average of 32.1 ED visits, middle income counties had an average of 38.0 ED visits,and high income counties had an average of 134 ED visits. The average % Bachelor's Degree or higher for low income counties was 17.1%, middle income counties was 18.7%, and high income counties was 25.4%. I calculated the mean % of the population that was a High School Graduate or higher for low income counties was 78.7%, middle income counties was 82.7%, and high income counties was 84.2%. The mean Unemployment rate by per capita income was 4.58% for low income counties, 4.36% for middle imcome counties, and 3.80% for high income counties. The mean % of Population in Poverty by per capita income was 18.7% for low income counties, 15.0% for middle income counties, and 11.7% for high income counties. The high income counties, as expected, had a higher average percents of their populations graduate high shcool, earn a Bachelor's degree, and number of ED visits than low income counties. They also had lower average umemployment and poverty rates than their counterparts.

The mean per capita income by  % of population with a Bachelor's Degree or higher was $39384, $40397, and $47779 while the mean median income was $48101, $50407, and $61735. As expected, higher percentages of bachelor degree holders is associated with higher incomes. The mean Unemployment by % of population with a Bachelor's Degree or higher is 4.66%, 4.4%, and 3.69%. The mean % of Population in Poverty by % of population with a Bachelor's Degree or higher is 16.9%, 16.0%, and 12.5%. As expected, higher percentages of bachelor degree holders is associated with lower rates of poverty and umemployment. In counties with a high proportion of bachelor degree holders, the trend that counties with higher per capita incomes having lower poverty rates still holds true.

## 4. Visualizing
```{r Visualizing , echo = FALSE}
ggplot(fulldata, aes(x=`% Bachelor s Degree or Higher`))+geom_point(aes(y=`Per Capita Income`, size=(fulldata$`Number of Opioid-Related Emergency Department Visits`)), shape = 21,colour='black',fill='#1876f0')+scale_size_continuous(name = "Opioid-Related ED Visits")+ ggtitle("Plot of Per Captia Income by % of a County's Population with a Bachelor's Degree or Higher")+ylab("Per Capita Income ($)") + xlab("County's Population with a Bachelor's Degree or Higher (%)")+scale_y_continuous(breaks=seq(0,80000,5000))+theme(axis.text.x = element_text(colour="#f08f18"),axis.text.y = element_text(colour="#65a11e"),plot.title = element_text(size=10))

fulldata1<-fulldata
fulldata1<-fulldata1%>% pivot_longer(contains("%"))

ggplot(fulldata1, aes(x = name, y = value, fill=value))+ geom_bar(stat="summary",fun.y="mean",fill=c("#27b3b6","#65a11e","#f08f18","#b6277f"))+theme(axis.text.x = element_text(angle=45, hjust=1, size=10))+ ggtitle("Education and Economic Statistics of Texas Counties")+ylab("Percentage of the Population (%)") + xlab("Average of All County Statistics")
```
The "Plot of Per Captia Income by % of a County's Population with a Bachelor's Degree or Higher" scatterplot shows a somewhat positive association between the proportion of a county's population with a Bachelor's degree or higher and an increasing per capita income which is too be expected since higher levels of education typically accompany higher paying jobs. I set the size of the points to be proportional to the number of opioid related ED visits. There is less of a clear trend when it comes to this variable but it does appear that a higher number of ED visits is associated with higher per capita income. The reason for this is equally obfuscated but it is possible that having excess income or higher SES may lead to more risky health behaviours. This may also be due to the fact that those with a higher SES and income are more likely to have access to and utilize healthcare.
For the bar graph there are not any relationships or trends to look at in particular. The average proportion of the population that has graduated high school or higher is above 80% which seems quite high although I do not have any standards to compare it too. The unemployment rate also seems quite low, looking it up right now, the US current unemployment rate is 3.7% which is below the average 4.25% rate for Texas counties. 
## 5. Dimensionality Reduction (20 pts) 
```{r Dimensionality Reduction 1}
#clean/scale data
fulldata<-fulldata%>%mutate("OpioidVisits"=cut(`Number of Opioid-Related Emergency Department Visits`, breaks=c(-Inf, 13,36,Inf), labels=c("low","med","high")))
fulldata_nums<-fulldata %>%select_if(is.numeric)%>%scale
rownames(fulldata_nums)<-fulldata$County
fulldata_pca<-princomp(fulldata_nums)
names(fulldata_pca)
#summarize results
summary(fulldata_pca, loadings=T)
#deciding how many PC's to keep
eigval<-fulldata_pca$sdev^2#square to convert SDs to eigenvalues
varprop=round(eigval/sum(eigval),2) #proportion of var explained by each PC
```

I scaled and selected the numeric variables from my data to use with princomp. I then used the summary function to look at the standard deviation, proportion of variance, and cummulative proportion of each PC. 
```{r Dimensionality Reduction 2}
ggplot()+geom_bar(aes(y=varprop,x=1:7),stat="identity")+xlab("")+geom_path(aes(y=varprop,x=1:7))+  geom_text(aes(x=1:7,y=varprop,label=round(varprop,2)),vjust=1,col="white",size=5)+  scale_y_continuous(breaks=seq(0,.6,.2),labels = scales::percent)+  scale_x_continuous(breaks=1:10)
```

The first 4 PC make up .78 (close to .8) of the total variance proportion and it's hard to tell where the "elbow" is so I chose them to continue with.
```{r Dimensionality Reduction 3}
library(ggrepel)
fulldata_df<- fulldata %>% mutate(PC1=fulldata_pca$scores[,1], PC2=fulldata_pca$scores[,2])
ggplot(fulldata_df, aes(PC1,PC2,color=OpioidVisits))+geom_point()+geom_text_repel(aes(label=County),color="black")
```

PC1 does not seem to adequately separate the dataset based on whether the county in question has a high, medium, or low amount of opioid related ED visits. PC2 also does not do a very good job of sepaparting the number opioid ED visits but a high number opioid ED visits does appear to be higher than the low and medium visits.
```{r Dimensionality Reduction 4}
fulldata_df<- fulldata %>% mutate(PC3=fulldata_pca$scores[,3], PC4=fulldata_pca$scores[,4])
ggplot(fulldata_df, aes(PC3,PC4,color=OpioidVisits))+geom_point()+geom_text_repel(aes(label=County),color="black")
```

Just like PC 1&2, PC4 does not seem to  separate the dataset based on whether the county in question has a high, medium, or low amount of opioid related ED visits nor does PC4. However,
```{r Dimensionality Reduction 5}
fulldata_pca$loadings[2:7,1:2]%>%as.data.frame%>%rownames_to_column %>% ggplot()+geom_hline(aes(yintercept=0),lty=2)+geom_vline(aes(xintercept=0),lty=2)+ylab("PC2")+xlab("PC1")+geom_segment(aes(x=0,y=0,xend=Comp.1,yend=Comp.2), arrow=arrow(),col="red")+geom_label(aes(x=Comp.1*1.1,y=Comp.2*1.1,label=rowname))
fulldata_pca$loadings[2:7,3:4]%>%as.data.frame%>%rownames_to_column %>% ggplot()+geom_hline(aes(yintercept=0),lty=2)+geom_vline(aes(xintercept=0),lty=2)+ylab("PC4")+xlab("PC3")+geom_segment(aes(x=0,y=0,xend=Comp.3,yend=Comp.4), arrow=arrow(),col="red")+geom_label(aes(x=Comp.3*1.1,y=Comp.4*1.1,label=rowname))
```
For the plot with PC1 and PC2: The angles between vectors interpreted as correlations so % having a bachelor's Degree or higher and Per capital income are highly correlated wiith each other and negatively correlated with PC1 and  % in poverty and unemployment rateare correlated with each other and psitively with PC1. % in poverty and unemployment rate are not strongly correlated with % having a bachelor's Degree or higher, Median household income, % high school graduate or higher, nor Per capital income.
For the plot with PC3 and PC4: % in poverty, % high school graduate or higher, and % having a bachelor's Degree or higher are highly correlated and ot highly correlated with Unemployment rate, median household income, and per capita income.
